prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-10-20 inst/MyArticle/paper.Rmd
rename tidy table to tidy data

```{r setup, include=FALSE}
## See ACS guidelines [https://publish.acs.org/publish/author_guidelines?coden=ancham#manuscript_text_components]
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 5)
ggplot2::theme_set(ggplot2::theme_classic())
library(BiocStyle)
options(width=70)
```

@@

We store all the data needed for analysis in a tidy table, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy table. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats, and enables the use of the methods without additional arguments. We show an example code for creating an MSFragger[@yu2020fast] configuration in the Appendix. For popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy tables. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (Figure \@ref(fig:tabCompletion).

@@

### Determining linear parameter combinations for treatment comparison

The package `r BiocStyle::Githubpkg("fgcz/prolfqua")` provides a function to determine the vector of $parameter$ weights $c$, from a linear models and a contrast specification string.

The linear model below explain the observed protein abundances using the explanatory variables `factor_1` and `factor_2` and the interaction among them `factor_1:factor_2`,

```{r}
Intensity ~ factor_1 + factor_2 + factor_1:factor_2
```

then the contrasts among `group_1` and `group_2` defined by `factor_1` can be specified using the string below.

```{r eval=FALSE, echo=TRUE}
c("contrast_name" = "factor_1group_1 - factor_1group_2")
```

Furthermore, contrasts for subgroups can be specified using the code below,

```{r eval=FALSE, echo=TRUE}
c("contrast_name" =
    "`factor_1group_1:factor_2group_a` - `factor_1group_1:factor_2group_b`")

@@

where `factor_x` is the name of the explanatory variable, and `group_x`  are group labels.

The following code shows an example where we specify two contrast: the first to compare Cells of type CMP/MEP with cells of type HSC, and the second to compare therapy NO with therapy NU for the celltype CMP/MEP [@meier2021reduced]. Finally, we compute the array of weights defining the contrast, used to multiply the model coefficient $\beta$.

```{r echo=TRUE}
Contrasts <- 
    c("CMP/MEPvsHSC" = "`CelltypeCMP/MEP` - `CelltypeHSC`",
   "NOvsHU" = "`class_therapyc.NO:CelltypeCMP/MEP` - `class_therapyp.HU:CelltypeCMP/MEP`")

m <- prolfqua::prolfqua_data('data_basicModel_p1807')
linfct <- prolfqua::linfct_from_model(m,as_list = FALSE)

prolfqua::linfct_matrix_contrasts(linfct, Contrasts )
```

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use a random subset of the Ionstar dataset containing $163$ proteins and $1258$ peptides. Peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:prepro) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%', fig.width = 8, fig.height = 8}
## load peptide level data
d <- prolfqua::prolfqua_data('data_ionstar')$filtered()
d <- prolfqua::LFQData$new(d$data, prolfqua::old2new(d$config))
## create R6 obejct
lfqd <- prolfqua::LFQData$new(d$data, d$config)

##  transform intensities
t <- lfqd$get_Transformer()

lfqd <-  t$log2()$robscale()$lfq
lfqd$rename_response("peptide_abundance")

## infer protein intensities from peptide intensity
agr <- lfqd$get_Aggregator()
lfqp <- agr$medpolish()
lfqp$rename_response("protein_abundance")

## plot panels A-D
pl <- lfqd$get_Plotter()
panelA <- pl$intensity_distribution_density() +
  ggplot2::labs(tag = "A") + ggplot2::theme(legend.position = "none")
panelB <- agr$plot()$plots[[1]] + ggplot2::labs(tag = "B")
panelC <- lfqp$get_Stats()$violin() + ggplot2::labs(tag = "C")
pl <- lfqp$get_Plotter()
panelD <- pl$boxplots()$boxplot[[1]] + ggplot2::labs(tag = "D")
ggpubr::ggarrange(panelA, panelB, panelC, panelD)
```

@@

## compute contrasts from linear model, moderate t-statistics and p-values
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
  prolfqua::ContrastsModerated$new()

# compute contrasts using imputation imputation and moderate
conI <- prolfqua::ContrastsSimpleImpute$new( lfqp, contrasts) |> 
  prolfqua::ContrastsModerated$new()

# merge contrasts, to obtain differences for all proteins
contrasts <- prolfqua::merge_contrasts_results(prefer = contr, add = conI)

## visualize results in panels
pl <- contrasts$merged$get_Plotter() 
panelA <- pl$histogram()$p.value + ggplot2::labs(tag = "A")                                                        
panelB <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom") +
   ggplot2::labs(tag = "B")

@@

```{r benchmarkROC}
ttt <- sapply(allBenchmarks, function(x){x$complete(FALSE)})
names(allBenchmarks)

for(i in seq_along(allBenchmarks)) {
  allBenchmarks[[i]]$pAUC()
}

res <- purrr::map_df(allBenchmarks, function(x){x$pAUC()})
res <- res |> dplyr::mutate(whatfix = dplyr::case_when(what == "scaled.beta.based.significance" ~ "scaled.p.value", TRUE ~ what))

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure 2 and Table 3). The contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. The developed framework is flexible enough to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables you to call various methods and quickly select the best differential expression analysis algorithm for your problem. It also can generate conclusive reports and can be used to benchmark MS software and statistical methods. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).

@@

>>>>

```{r setup, include=FALSE}
## See ACS guidelines [https://publish.acs.org/publish/author_guidelines?coden=ancham#manuscript_text_components]
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      fig.width = 7,
                      fig.height = 5)
ggplot2::theme_set(ggplot2::theme_classic())
library(BiocStyle)
options(width = 70)
```

@@

We store all the data needed for analysis in a data frame as tidy data, i.e., every column is a variable, every row is an observation, and every cell is a single value [@tidydata2014]. Using an R6 [@R6cite] configuration object (Figure \@ref(fig:LFQData)), we specify what variable is in which column, making it easy to integrate new inputs in `r BiocStyle::Githubpkg("fgcz/prolfqua")` if provided as a tidy data. For example, to visualize tidy Spectronaut[@bruderer2015extending], DiaNN[@demichev2020dia], or Skyline[@maclean2010skyline] outputs, or data in `r BiocStyle::Biocpkg("MSstats")`[@Choi2014] format, only a few lines of code are needed to update the `r BiocStyle::Githubpkg("fgcz/prolfqua")` `AnalysisTableConfiguration` configuration. The configuration encapsulates the differences in column names among the various input formats, and enables the use of the methods without additional arguments. We show an example code for creating an MSFragger[@yu2020fast] configuration in the Appendix. For popular software like MaxQuant[@Cox2008MaxQuant], or MSFragger, which stores the same variable, e.g., intensity, in multiple columns, one for each sample, we implemented methods that transform the data into tidy data. Relying on the tidy data table enables us to interface with many data manipulation, visualization, and modeling methods, implemented in base R [@RCoreTeam2021] and the tidyverse [@tidyverse2019], easily. We use R6 classes to structure the functionality of the package (see Figure \@ref(fig:LFQData) and Figure \@ref(fig:ContrastUML)). R6 classes are well supported by command-line completion features in RStudio [@RStudio2022], and help to implement argument free functions (Figure \@ref(fig:tabCompletion).

@@

The package `r BiocStyle::Githubpkg("fgcz/prolfqua")` provides functions to determine the vector of $parameter$ weights $c$, from a linear model and a contrast specification string.

The linear model below explain the observed protein abundances using the explanatory variables `condition_` and `batch_` and the interaction among them `condition_:batch_`,

```{r, echo = TRUE}
data_Yeast2Factor <- prolfqua::prolfqua_data("data_Yeast2Factor")
mlm <- prolfqua::strategy_lm(transformedIntensity ~ condition_ + batch_ + condition_:batch_)
mm <- prolfqua::build_model(data_Yeast2Factor$data, mlm, subject = "protein_Id")
```

then the contrasts among the groups `Ethanol` and `Glucose` defined by the factor `condition`, and similarly for the difference of the groups defined by `batch_`, can be specified as:

```{r contrast1,  echo=TRUE}
d1 <- c("diff_condition" = "condition_Ethanol - condition_Glucose",
  "diff_batch" = "batch_p2370 - batch_p2691")
```

Furthermore, the difference among `Ethanol` and `Glucose` within the batches `p2370` and `p2691` can be specified as follows:

```{r contrast2, eval=TRUE, echo=TRUE}
d2 <- c("difference_within_p2370" =
    "`condition_Glucose:batch_p2370` - `condition_Ethanol:batch_p2691`",
  "difference_within_p2691" = 
   "`condition_Glucose:batch_p2691` - `condition_Ethanol:batch_p2691`")
```

Finally, we can examine if the differences among `conditions_` observed in both batches significantly differ by specifying the contrast:

```{r contrast3, eval=TRUE, echo=TRUE}
d3 <- c("difference_of_difference" = "difference_within_p2370 - difference_within_p2691")
```

The code to compute the array of weights $c$ used to multiply the model coefficient $\beta$, to obtain the effect size  estimates, is shown below:

```{r contrast4, echo=TRUE}
Contrasts <- c(d1, d2, d3)
linfct <- prolfqua::linfct_from_model(mm$modelDF$linear_model[[1]],
                                      as_list = FALSE)
mC <- prolfqua::linfct_matrix_contrasts(linfct, Contrasts)
```

while Table \@ref(tab:contrcoefficient) shows the weight $c$ of each model parameter in each contrast.

```{r contrcoefficient}
knitr::kable(mC, caption = "Weights of each model parameter.",format = "latex")

@@

This example is taken from the `r BiocStyle::Githubpkg("fgcz/prolfqua")` package vignette file.
Given the coefficients, $c$ contrasts for models obtained with `r BiocStyle::Biocpkg('proDA')`, or `r BiocStyle::Biocpkg('msqrob2')` can be determined, enabling us to integrate these packages transparently.

@@

The code snippets in this section demonstrate how a differential expression analysis workflow can be implemented using the `r BiocStyle::Githubpkg("fgcz/prolfqua")` R package. To speed up the computation of these examples, we use subset of the Ionstar dataset generated by randomly selecting $400$ proteins. Peptide abundances are $\log_2$ transformed and robust z-score scaled using the method `robscale`. Using the `LFQDataPlotter` class, we show the distribution of the normalized peptide abundances in Figure \@ref(fig:prepro) Panel A. Afterwards, protein intensities are estimated from peptide intensities using Tukey's median polish. Figure \@ref(fig:prepro) Panel B shows the peptide intensities and the estimated protein intensities. Next, we compute the standard deviation of all the proteins in each group and display their distribution using violin plots (Panel C). Finally, we create a boxplot (Panel D) showing the abundance of one protein.

```{r prepro,  echo = TRUE , fig.cap="(ref:scaling)", out.width = '90%', fig.width = 8, fig.height = 8}
R.version.string; packageVersion("prolfqua")

# read MQ peptide.txt and annotation table
startdata <- prolfqua::tidyMQ_Peptides(system.file(
  "samples/maxquant_txt/tiny2.zip", package = "prolfqua"))
annot <- readxl::read_xlsx(system.file(
  "samples/maxquant_txt/annotation_Ionstar2018_PXD003881.xlsx", package = "prolfqua"))
startdata <- dplyr::inner_join(annot, startdata, by = "raw.file")
# create proflqua configuration
config <- prolfqua::create_config_MQ_peptide()
config$table$factors[['dilution.']] = "sample"
## create R6 object
lfqd <- prolfqua::LFQData$new(startdata, config, setup = TRUE) 
lfqd$hierarchy_counts()
## remove observation with 0 intensity and filter for 2 peptides per protein
lfqd$remove_small_intensities()
lfqd$filter_proteins_by_peptide_count()
lfqd$hierarchy_counts()
##  transform intensities
lfqd <- lfqd$get_Transformer()$log2()$robscale()$lfq
lfqd$rename_response("log_peptide_abundance")
agr <- lfqd$get_Aggregator()
lfqp <- agr$medpolish()
lfqp$rename_response("log_protein_abundance")
## plot panels A-D
pl <- lfqd$get_Plotter()
panelA <- pl$intensity_distribution_density() +
  ggplot2::labs(tag = "A") + ggplot2::theme(legend.position = "none")
panelB <- agr$plot()$plots[[54]] + ggplot2::labs(tag = "B")
panelC <- lfqp$get_Stats()$violin() + ggplot2::labs(tag = "C")
pl <- lfqp$get_Plotter()
panelD <- pl$boxplots()$boxplot[[54]] + ggplot2::labs(tag = "D")
ggpubr::ggarrange(panelA, panelB, panelC, panelD)

@@

## compute contrasts from linear model
contr <- prolfqua::Contrasts$new(models, contrasts) 
# compute contrasts using imputation imputation
conI <- prolfqua::ContrastsSimpleImpute$new( lfqp, contrasts)
# merge contrasts, to obtain differences for all proteins
contrasts <- prolfqua::merge_contrasts_results(prefer = contr, add = conI)
# moderate t-statistics and p-values
prolfqua::ContrastsModerated$undebug("get_contrasts")
contrasts <-  contrasts$merged |> prolfqua::ContrastsModerated$new() 
## visualize results in panels
pl <- contrasts$get_Plotter() 
panelA <- pl$histogram()$p.value + ggplot2::labs(tag = "A")                                                        
panelB <- pl$volcano()$FDR +
   ggplot2::theme(legend.position = "bottom") +
   ggplot2::labs(tag = "B")

@@

```{r benchmarkROC}
ttt <- sapply(allBenchmarks, function(x){x$complete(FALSE)})
for (i in seq_along(allBenchmarks)) {
  allBenchmarks[[i]]$pAUC()
}
res <- purrr::map_df(allBenchmarks, function(x){x$pAUC()})
res <- res |> dplyr::mutate(whatfix = dplyr::case_when(what == "scaled.beta.based.significance" ~ "scaled.p.value", TRUE ~ what))

@@

`r BiocStyle::Githubpkg("fgcz/prolfqua")` is an feature-rich, object oriented R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. While other R packages for differential expression analysis of proteins typically only implement one modeling approach, `r BiocStyle::Githubpkg("fgcz/prolfqua")` supports various types of models (see Figure 2 and Table 3). The contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for adding new features, e.g., generalized linear models to model the presence or absence information of a protein, or robust linear models, in the future. Furthermore, the developed framework is flexible enough to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables you to call various methods and quickly select the best differential expression analysis algorithm for your problem. It also can generate conclusive reports and can be used to benchmark MS software and statistical methods. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).

@@

```{r roctest}
ttmsqrob <- allBenchmarks$benchmark_msqrob$data()
ddmsqrob <- pROC::roc(ttmsqrob$TP, ttmsqrob$scaled.p.value, partial.auc = c(1, 0.9))
ttprolfqua <- allBenchmarks$benchmark_merged$data()
ddprolfqua <- pROC::roc(ttprolfqua$TP, ttprolfqua$scaled.p.value, partial.auc = c(1, 0.9))
ttproda <- allBenchmarks$benchmark_proDA$data()
ddproda <- pROC::roc(ttproda$TP, ttproda$scaled.p.value, partial.auc = c(1, 0.9))

msqrob_vs_prolfqua <- pROC::roc.test(ddmsqrob,ddprolfqua)
msqrob_vs_proda <- pROC::roc.test(ddmsqrob,ddproda)
prolfqua_vs_proda <- pROC::roc.test(ddprolfqua,ddproda)
